Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland

### 2017-09-04 paper/main.tex
included Walids suggestions

We isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI).
For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (Supplementary Figure 1, \autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility between the biological replicates (Supplementary Figure 2,3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 4a).
In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15) (\autoref{F1}d).\\
To further characterise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
As shown in Table 1 multiple clusters of cells have been assigned to the same putative cell type.
We found that in these instances the cells showed a high similarity and expressed the same marker genes but derived from different developmental stages, suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\

In the luminal compartment we found two larger subgroups (\autoref{F2}a).
C1-C5 showed characteristics of hormone-sensing cells (Hs) such as high expression levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c).
Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g. \textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
C8-C11 expressed expressed milk proteins (e.g. \textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).
We also found C6/7 which were devoid of hormone-receptors and milk proteins but expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).\\

@@

Our data illustrates how the cellular composition of the gland changes during the pregnancy cycle.
The luminal compartment shifts from giving rise to mainly hormone sensing cells to producing alveolar, milk-producing cells during pregnancy and lactation. 
The basal compartment on the other hand differentiates to produce oxytocin-sensing, myoepithelial cells that enable duct contraction and milk secretion during lactation.
At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}c).
Here, the effect of parity is most pronounced for the luminal progenitor cell types (\autoref{F1}c, Supplementary Figure 7).
C6 and C7 both showed characteristics of luminal progenitor cells, but we identified C6 practically only in the nulliparous gland and C7 in the gland from parous animals.
To ensure that C7 still represents the progenitor population in the post-parous gland, we identified genes that distinguish C6 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C7.
Indeed, we find genes that are differentially expressed between C6 and the rest of the luminal compartment to show the same trend between C7 and the luminal compartment (\autoref{F5}a).
In a similar manner we can distinguish C7 from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 7b and c).
From this we conclude, that C7 indeed represents the post-parity luminal progenitor population.
Interestingly, genes that were up-regulated in C7 compared to C6 were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F5}b, Supplementary Fig 7d).
These included genes that play roles in various processes during lactation, e.g. milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea})\autocite{Wang2012}.
Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment.
Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}\autocite{Clarkson2003}.
The up-regulation of these genesets was also present in C1 (Hsp) and in some cases even the differentiated C3 (Hsd) cluster, yet the effect remained strongest in the progenitor cells (\autoref{F5}c-d).
Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
Interestingly, C1 and C3 cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the C7 cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
This is in agreement with the overexpression of genes related to the production and secretion of milk.
Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
This memory could potentially prime progenitor cells towards the alveolar fate to facilitate alveologenesis in subsequent pregnancies.

@@

We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage.
This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
We note, however, that only some of the clusters can be fully characterized by a single marker gene.
Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3}.
This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
We do note, however, that cells form C15 (Prc) expresses low levels of high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
In addition, cells from the C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
Lp gives rise to intermediate states of either hormone-sensing or secretory cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells.
Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
Here we identified genes that are lineage-specific, thus enabling us to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
We assume that C7 overlaps at least partially with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.

@@

%---------------------------------------------Figure 1--------------------------------------------------------------
\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
\fakefigure{F1}
\textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
\textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
Cells are coloured by the four developmental time-points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
\textbf{(c)} Same as \textbf{(b)} but colouring cells by clusters. The smaller t-SNEs are coloured by the normalised log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
%-------------------------------------------------------------------------------------------------------------------

%---------------------------------------------Figure 2--------------------------------------------------------------
\textbf{Figure 2. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
\fakefigure{F2}
\textbf{(a)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
\textbf{(b)} t-SNE with overlaid expression of cluster specific genes.
\textbf{(b)} Heatmap highlighting key marker genes that were used to infer putative identities.
Colour scale represents log-transformed and normalised UMI counts scaled to a maximum of 1 per row.
The upper panel shows genes that were used to distinguish between luminal and basal cells.
Upper bars represent the cluster assignment and stages for the individual cells.
For visualization purposes only 100 randomly selected cells were shown for large clusters.
%-------------------------------------------------------------------------------------------------------------------

>>>>

We isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI).
For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (Supplementary Figure 1, \autoref{F1}a).
All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility between the biological replicates (Supplementary Figure 2,3).
Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
This resulted in a coarse clustering into 13 groups (Supplementary Figure 4a).
In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d).\\
To further characterise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
As shown in Table 1 multiple clusters of cells have been assigned to the same putative cell type.
We found that in these instances the cells showed a high similarity and expressed the same marker genes but derived from different developmental stages, suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\

In the luminal compartment we found two larger subgroups (\autoref{F2}a).
C1-C5 showed characteristics of hormone-sensing cells (Hs) such as high expression levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c).
Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g. \textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).\\
In contrast, C8-C11 expressed expressed milk proteins (e.g. \textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).

@@

Our data illustrates how the cellular composition of the gland changes during the pregnancy cycle.
The luminal compartment shifts from giving rise to mainly hormone sensing cells to producing alveolar, milk-producing cells during pregnancy and lactation. 
The basal compartment on the other hand differentiates to produce oxytocin-sensing, myoepithelial cells that enable duct contraction and milk secretion during lactation.
At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}c).
Here, the effect of parity is most pronounced for the luminal progenitor cell types (\autoref{F1}c, Supplementary Figure 7).
C6 and C7 both showed characteristics of luminal progenitor cells, but we identified C6 practically only in the nulliparous gland and C7 in the gland from parous animals.
To ensure that C7 still represents the progenitor population in the post-parous gland, we identified genes that distinguish C6 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C7.
Indeed, we find genes that are differentially expressed between C6 and the rest of the luminal compartment to show the same trend between C7 and the luminal compartment (\autoref{F5}a).
In a similar manner we can distinguish C7 from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 7b and c).
From this we conclude, that C7 indeed represents the post-parity luminal progenitor population.
Interestingly, genes that were up-regulated in C7 compared to C6 were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F5}b, Supplementary Fig 7d).
These included genes that play roles in various processes during lactation, e.g. milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea})\autocite{Wang2012}.
Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment.
Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}\autocite{Clarkson2003}.
The up-regulation of these genesets was also present in C1 (Hsp) and in some cases even the differentiated C3 (Hsd) cluster, yet the effect remained strongest in the progenitor cells (\autoref{F5}c-d).
Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
Interestingly, C1 and C3 generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the C7 cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
This is in agreement with the overexpression of genes related to the production and secretion of milk.
Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
This memory could potentially prime progenitor cells towards the alveolar fate to facilitate alveologenesis in subsequent pregnancies.

@@

We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage.
This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
We note, however, that only some of the clusters can be fully characterized by a single marker gene.
Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3} and \autoref{F4}.
This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
We do note, however, that cells form C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
In addition, cells from the C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
Lp gives rise to intermediate states of either hormone-sensing or secretory cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells.
Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
Here we identified genes that are lineage-specific, thus enabling us to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
We assume that C7 overlaps at least partially with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.

@@

%---------------------------------------------Figure 1--------------------------------------------------------------
\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
\fakefigure{F1}
\textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
\textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
Cells are coloured by the four developmental time-points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
\textbf{(c)} Same as \textbf{(b)} but colouring cells by clusters.
\textbf{(d)} t-SNEs coloured by the normalised log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
%-------------------------------------------------------------------------------------------------------------------

%---------------------------------------------Figure 2--------------------------------------------------------------
\textbf{Figure 2. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
\fakefigure{F2}
\textbf{(a)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
\textbf{(b)} t-SNEs with overlaid expression of cluster specific genes.
\textbf{(c)} Heatmap highlighting key marker genes that were used to infer putative identities.
Colour scale represents log-transformed and normalised UMI counts scaled to a maximum of 1 per row.
The upper panel shows genes that were used to distinguish between luminal and basal cells.
Upper bars represent the cluster assignment and stages for the individual cells.
For visualization purposes only 100 randomly selected cells were shown for large clusters.
%-------------------------------------------------------------------------------------------------------------------
